---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r}
library(plyr)
detach("package:plyr", unload=TRUE)
library(dplyr)
library(ggplot2)
library(timeDate)
library(lattice)
activity <- read.csv("activity.csv")
activity <- tbl_df(activity)
activity
```
## Mean total number of steps taken per day
### total number of steps taken per day
```{r}
by_date = group_by(activity, date)
steps_per_day = summarize(by_date, totalSteps = sum(steps, na.rm = TRUE), avgSteps = mean(steps,na.rm = TRUE), mediSteps = median(steps, na.rm = TRUE))
steps_per_day
```
### Histogram of the total number of steps taken each day

```{r}
hist(steps_per_day$totalSteps)
```
### Mean and median number of steps taken each day?

```{r}
select(steps_per_day,date, avgSteps, mediSteps)
```
## Average daily activity pattern
### Time Series plot of 5 minute interval 
```{r}
by_interval = group_by(activity, interval)
steps_per_interval = summarize(by_interval, totalSteps = sum(steps, na.rm = TRUE), avgSteps = mean(steps,na.rm = TRUE), mediSteps = median(steps, na.rm = TRUE))
steps_per_interval
ggplot(steps_per_interval, aes(x = factor(interval), y = avgSteps)) + geom_line(aes(group =1))
```
### The interval with the maximum avg no. of Steps

835 

## Imputing missing values

### Count number of missing values

```{r}
count(activity[is.na(activity),])
```

### Strategy for imputting missing values
To account for missing values, i am using the average value from the same interval to populate values to the missing elemnets. To do this;

* Merge my table with missing values, with the summarized table created earlier, using interval as the intersection. This creates a new table with the original table colums plus the value of the average of each corresponding interval.
* Next I check the NAs in the steps column and replace them with the average steps column value.
*Create a new dataset that is equal to the original dataset but with the missing data filled in

```{r}
activity2 = tbl_df(merge(activity,steps_per_interval, by = "interval", all.x =TRUE))
activity_imputed = tbl_df(transform(activity2, steps = ifelse(is.na(activity2$steps),activity2$avgSteps,activity2$steps)))
activity_imputed = select(activity_imputed, steps, date,interval)
```

### New dataset
```{r}
activity_imputed
```
### Histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day for new data set

```{r}
by_date2 = group_by(activity_imputed, date)
steps_per_day2 = summarize(by_date2, totalSteps = sum(steps, na.rm = TRUE), avgSteps = mean(steps,na.rm = TRUE), mediSteps = median(steps, na.rm = TRUE))
steps_per_day2
hist(steps_per_day2$totalSteps)
```


## Differences in activity patterns between weekdays and weekends?

### new factor variable in the dataset with two levels – “weekday” and “weekend”

```{r}
activity_imputed = tbl_df(transform(activity_imputed, weekday = ifelse(isWeekday(activity_imputed$date), "weekday", "weekend")))
```

### Panel Plot

```{r}
by_interval_weekday = group_by(activity_imputed, interval, weekday)
steps_by_interval_weekday = summarize(by_interval_weekday, avgSteps = mean(steps))
ggplot(steps_by_interval_weekday, aes(interval, avgSteps)) + geom_line() + facet_wrap(~ weekday, ncol=1)

```

